Finding a closest point in a lattice of Voronoi's first kind

### 2014-01-09 paper.tex
minor.  Need to cleanup this notation

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$, let $S$ be a subset of the indices of $p$, and let $B$ be the matrix from~\eqref{eq:matrxBobtusebasis} with columns given by the superbasis vectors $b_1,\dots,b_{n+1}$.  The following equalities hold:
\begin{enumerate}
\item  ${\displaystyle \|Bp\|^2 - \|B(p - I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i)}$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsinc}
%\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S + I_T)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsincanddec}
\end{enumerate}
where $q_{ij} = \dotprod{b_i}{b_j} \leq 0$ are the Selling parameters.
\end{lemma}
\begin{proof}
We give a proof for part~\ref{eq:lem:decSellingsdec}.  The proof for part~\ref{eq:lem:decSellingsinc} is similar.  Observe that
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
\]
and
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
\]
since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
\begin{align*}
\|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
&= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
&= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
\end{align*}
This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - I_S$.  It follows that
\[
\|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
\]
where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
\[
d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
\]
If $i \in S$ and $j \notin S$ then
\[
d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
\]
while if $i \notin S$ and $j \in S$ then
\[
d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
\]
We thus have
\begin{align*}
\|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
\end{align*}
where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

@@

\begin{lemma}\label{lem:rngdecreases}
Let $Bu$ with $u \in \ints^{n+1}$ be a lattice point that is $\ell$-close to $y = Bz$ where $\ell > 0$.  Let $g \in \{0,1\}^{n+1}$ be such that
\begin{equation}\label{eq:qismin}
\|B(z - u - g)\|^2 = \min_{p \in \{0,1\}^{n+1}}\|B(z - u - p)\|^2.
\end{equation}
The lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
\end{lemma}
\begin{proof}
Because $Bu$ is $\ell$-close to $y$ there exists $v \in \ints^{n+1}$ with $\rng(v) = \ell$ such that $B(u+v)$ is a closest lattice point to $y$.  Let $A$ be the (possibly empty) subset of indices $\{1,\dots,n+1\}$ such that $g_i = 1$ and $v_i = \min(v)$ and let $D$ be the (possibly empty) subset of indices such that $g_i = 0$ and $v_i = \max(v)$. That is,
\[
A = \{i \mid g_i = 1, v_i = \min(v) \}, \qquad D = \{i \mid g_i = 0, v_i = \max(v) \}.
\]
Put $d = g - I_A + I_D$ and put $w = v - d$.  Observe that $d_i = 0$ whenever $v_i = \min(v)$ and so $\min(w) = \min(v-d) = \min(v)$.  Also, $d_i = 1$ whenever $v_i = \max(v)$ and so $\max(w) = \max(v-d) = \max(v) - 1$.  Thus,
\[
\rng(w) = \max(v) - 1 - \min(v) = \rng(v) - 1 = \ell - 1.
\]
The lemma will follow if we show that $B(u+g+w)$ is a closest lattice point to $y$ since then the lattice point $B(u+g)$ with be $(\ell-1)$-close to $y$.  We show this in two stages.  First put $h = v - g + I_A$.  Suppose that $B(u+g+h)$ is not a closest lattice point to $y = Bz$.  That is, suppose that 
\[
\|B(z-u-g-h)\|^2 = \|B(z-u-v-I_A)\|^2 > \|B(z-u-v)\|^2.
\]

In this case, Lemma~\ref{lem:vqcontramin} (proved below) with $t = z-u$ and $S = A$ shows that 
\begin{equation}\label{eq:lem:rng:contag}
\|B(z- u - g)\|^2 > \min_{p \in \{0,1\}^{n+1}}\|B(z- u - p)\|^2,
\end{equation}
contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+h)$ is a closest lattice point to $y$.  Now suppose that $B(u+g+w) = B(u+g+h-I_D)$ is not a closest lattice point to $y$.  That is, suppose that
\[
\|B(z-u-g-h+I_D)\|^2 > \|B(z-u-g-h)\|^2.
\]
Lemma~\ref{lem:vqcontramax} (proved below) with $t = z-u$ and $S = D$ shows that~\eqref{eq:lem:rng:contag} occurs again, contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+w)$ is a closest lattice point to $y$ and, since $\rng(w) = \ell-1$, the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}

>>>>

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$, let $S$ and $T$ be subsets of the indices of $p$, and let $B$ be the matrix from~\eqref{eq:matrxBobtusebasis} with columns given by the superbasis vectors $b_1,\dots,b_{n+1}$.  The following equalities hold:
\begin{enumerate}
\item  ${\displaystyle \|Bp\|^2 - \|B(p - I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i)}$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsinc}
% \item  \label{eq:lem:decSellingsincanddec} \begin{align*}
% \|Bp\|^2 - \|B(p + I_S + I_T)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2p_j - 2p_i) + \sum_{i\in S}\sum_{j\in S} q_{ij}, 
% \end{align*}
\end{enumerate}
where $q_{ij} = \dotprod{b_i}{b_j} \leq 0$ are the Selling parameters.
\end{lemma}
\begin{proof}
We give a proof for part~\ref{eq:lem:decSellingsdec}.  The proof for part~\ref{eq:lem:decSellingsinc} is similar.  Observe that
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
\]
and
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
\]
since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
\begin{align*}
\|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
&= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
&= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
\end{align*}
This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - I_S$.  It follows that
\[
\|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
\]
where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
\[
d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
\]
If $i \in S$ and $j \notin S$ then
\[
d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
\]
while if $i \notin S$ and $j \in S$ then
\[
d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
\]
We thus have
\begin{align*}
\|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
\end{align*}
where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

@@

\begin{lemma}\label{lem:rngdecreases}
Let $Bu$ with $u \in \ints^{n+1}$ be a lattice point that is $\ell$-close to $y = Bz$ where $\ell > 0$.  Let $g \in \{0,1\}^{n+1}$ be such that
\begin{equation}\label{eq:qismin}
\|B(z - u - g)\|^2 = \min_{p \in \{0,1\}^{n+1}}\|B(z - u - p)\|^2.
\end{equation}
The lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
\end{lemma}
\begin{proof}
Because $Bu$ is $\ell$-close to $y$ there exists $v \in \ints^{n+1}$ with $\rng(v) = \ell$ such that $B(u+v)$ is a closest lattice point to $y$.  Define subsets of indices
\[
A = \{i \mid g_i = 1, v_i = \min(v) \}, \qquad D = \{i \mid g_i = 0, v_i = \max(v) \}
\]
and put $d = g - I_A + I_D \in \{0,1\}^{n+1}$ and $w = v - d$.  Observe that $d_i = 0$ whenever $v_i = \min(v)$ and so $\min(w) = \min(v-d) = \min(v)$.  Also, $d_i = 1$ whenever $v_i = \max(v)$ and so $\max(w) = \max(v-d) = \max(v) - 1$.  Thus,
\[
\rng(w) = \max(v) - 1 - \min(v) = \rng(v) - 1 = \ell - 1.
\]
The lemma will follow if we show that $B(u+g+w)$ is a closest lattice point to $y$ since then the lattice point $B(u+g)$ with be $(\ell-1)$-close to $y$.  We show this in two stages.  First put $f = g - I_A \in \{0,1\}^{n+1}$ and  $h = v - f = v - g + I_A$.  Suppose that $B(u+g+h)$ is not a closest lattice point to $y = Bz$.  That is, with $t = z-u$, suppose that
\[
\|B(t-g-h)\|^2 = \|B(t-v-I_A)\|^2 > \|B(t-v)\|^2.
\]
By part~\ref{eq:lem:decSellingsdec} of Lemma~\ref{lem:decSellings} we obtain the inequality
\[
\|B(t-v)\|^2 - \|B(t-v-I_A)\|^2 =  \sum_{i \in A}\sum_{j \notin A}q_{ij}(1 + 2(t_j-t_i) + 2(v_i-v_j)) < 0.
\]
Because $v_i = \min(v)$ only when $f_i= g_i - 1 = 0$ and because $v_i = \min(v)$ for $i \in A$ we have $f_i - f_j \geq v_i - v_j$ for all $i \in S$ and $j =1,\dots,n+1$.  Now, since $q_{ij} \leq 0$ for $i\neq j$,
\begin{align*}
\|B(z-u-f)\|^2 - \|B(z-u-g)\|^2 &= \|B(t-f)\|^2 - \|B(t-f-I_A)\|^2 \\
&= \sum_{i \in A}\sum_{j \notin A}q_{ij}(1 + 2(t_j-t_i) + 2(f_i-f_j)) \\
&\leq  \sum_{i \in A}\sum_{j \notin A}q_{ij}(1 + 2(t_j-t_i) + 2(v_i-v_j)) < 0,
\end{align*}
contradicting~\eqref{eq:qismin} because $f \in \{0,1\}^{n+1}$.  Thus, $B(u+g+h)$ is a closest lattice point to $y = Bz$.

Now suppose that $B(u+g+w) = B(u+g+h-I_D)$ is not a closest lattice point to $y$.  That is, suppose that
\[
\|B(t-g-w)\|^2 = \|B(t-g-h+I_D)\|^2 > \|B(t-g-h)\|^2.
\]
By part~\ref{eq:lem:decSellingsinc} of Lemma~\ref{lem:decSellings} we obtain the inequality
\[
\|B(t-v)\|^2 - \|B(t-v-I_A)\|^2 =  \sum_{i \in A}\sum_{j \notin A}q_{ij}(1 + 2(t_j-t_i) + 2(v_i-v_j)) < 0.
\]

% In this case, Lemma~\ref{lem:vqcontramin} (proved below) with $t = z-u$ and $S = A$ shows that 
% \begin{equation}\label{eq:lem:rng:contag}
% \|B(z- u - g)\|^2 > \min_{p \in \{0,1\}^{n+1}}\|B(z- u - p)\|^2,
% \end{equation}
% contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+h)$ is a closest lattice point to $y$.  Now suppose that $B(u+g+w) = B(u+g+h-I_D)$ is not a closest lattice point to $y$.  That is, suppose that
% \[
% \|B(z-u-g-h+I_D)\|^2 > \|B(z-u-g-h)\|^2.
% \]
% Lemma~\ref{lem:vqcontramax} (proved below) with $t = z-u$ and $S = D$ shows that~\eqref{eq:lem:rng:contag} occurs again, contradicting~\eqref{eq:qismin}.

Thus, $B(u+g+w)$ is a closest lattice point to $y$ and, since $\rng(w) = \ell-1$, the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}
